---
title: "Versioning large objects with zudedup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Versioning large objects with zudedup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(zudedup)
```

Keeping every version of a large object costs the size of every version,
unless the versions share their storage. zudedup makes them share it: it
splits each version into chunks whose boundaries depend only on the bytes
around them, names each chunk by its hash, and stores each chunk once.

## Two versions of a data frame

Take a data frame, and a second version with a few rows changed.

```{r}
set.seed(1)
v1 <- data.frame(
  id = 1:50000,
  group = sample(letters, 50000, replace = TRUE),
  value = round(rnorm(50000), 4)
)
v2 <- v1
v2$value[c(100, 25000, 49000)] <- 0
x1 <- serialize(v1, NULL, xdr = FALSE)
x2 <- serialize(v2, NULL, xdr = FALSE)
length(x1)
```

`dedup_chunk()` shows the chunks. Their sizes cluster around `avg`, 8 KiB
by default, and no chunk is longer than `max`.

```{r}
chunks <- dedup_chunk(x1)
nrow(chunks)
summary(chunks$length)
```

A manifest records the chunk digests in order, with the size and the
digest of the whole object. Comparing two manifests says how much the
versions share:

```{r}
m1 <- dedup_manifest(x1)
m2 <- dedup_manifest(x2)
m1
d <- dedup_diff(m1, m2)
d$ratio
length(d$added)
```

Changing three values touched three chunks; the rest of the second version
is already there.

## A store

A store keeps each chunk once. Putting the first version writes all its
chunks; putting the second writes only the ones it does not share.

```{r}
store <- dedup_store(file.path(tempdir(), "vignette-store"), create = TRUE)
store
dedup_missing(store, m1) |> length()
dedup_put(store, x1)
dedup_missing(store, m2) |> length()
dedup_put(store, x2)
length(dedup_missing(store, m2))
```

`dedup_get()` reassembles a version from its manifest, checking every
chunk against its digest, so a store cannot hand back the wrong bytes.

```{r}
identical(unserialize(dedup_get(store, m2)), v2)
```

The manifest is the only way back to an object: keep it, for instance in a
database or next to a name. When a version is no longer needed, pass the
manifests you keep to `dedup_gc()`, which deletes every chunk none of them
uses.

```{r}
deleted <- dedup_gc(store, keep = m2)
length(deleted)
identical(unserialize(dedup_get(store, m2)), v2)
```

## Choosing the hash

The default hash, XXH3-128, is fast and is not cryptographic: someone who
can choose what goes into a store could make two chunks with the same
digest. That does not matter for a cache or a private store, and it does
for a store shared with parties you do not trust. There, use
`hash = "sha256"`, which needs the zucrypt package:

```{r, eval = requireNamespace("zucrypt", quietly = TRUE)}
dedup_manifest(x1, hash = "sha256")
```

## Other backends

`dedup_store()` keeps chunks in a directory. `dedup_backend()` builds a
store from five functions of your own, so chunks can live anywhere: here,
an environment.

```{r}
env <- new.env()
mem <- dedup_backend(
  has = function(hashes) vapply(hashes, exists, TRUE, envir = env,
                                inherits = FALSE, USE.NAMES = FALSE),
  get = function(hash) get0(hash, envir = env, inherits = FALSE),
  put = function(hash, bytes) assign(hash, bytes, envir = env),
  delete = function(hashes) rm(list = hashes, envir = env),
  list = function() ls(env)
)
m <- dedup_put(mem, x1)
length(ls(env))
identical(dedup_get(mem, m), x1)
```

A backend's `put` must be atomic: once it returns, the chunk is there
whole, or `has()` must not report it.

```{r, include = FALSE}
unlink(file.path(tempdir(), "vignette-store"), recursive = TRUE)
```
